\addcontentsline{toc}{chapter}{Conclusions}
\chapter*{Conclusions}

The detailed analysis and implementation work conducted in this project have confirmed that while the digital manufacturing industry has advanced significantly in terms of hardware, the software infrastructure for machine control remains rooted in the mid-20th century. The reliance on G-code creates a significant barrier to efficiency, customization, and safety.

The Problem-Based Learning (PBL) investigation highlighted three critical issues: the lack of readability and modifiability of machine instructions, the absence of abstraction which hides design intent, and the steep learning curve required to debug or optimize manufacturing processes manually. The proposed solution—the GG-Code DSL—emerges as a necessary and viable innovation.

By introducing high-level programming concepts such as variables, loops, functions, and semantic operations into the manufacturing workflow, the DSL transforms machine instructions from cryptic data streams into readable, maintainable software code. The successful specification of the formal grammar (Chapter 2) and the implementation of a robust parser (Chapter 3) demonstrate the technical feasibility of this approach. The parser is now capable of interpreting complex motion commands, geometric primitives, and algorithmic logic, validating the core architectural decisions.

In conclusion, the project has progressed from problem definition to a complete, working compiler. Beyond the language frontend, the semantic analyzer and code generator translate the validated Abstract Syntax Tree into standard G-code, and the language has been extended into a strict superset of G-code that can ingest, edit, and losslessly round-trip existing machine programs. The layer- and height-anchored constructs, together with a working \texttt{Pause} command and a two-pass injection model, let users apply precise, declarative modifications to real slicer output --- demonstrated on a production file of over 140{,}000 lines that compiles in a fraction of a second while preserving every machine instruction. A web-based IDE with a three-dimensional toolpath visualizer, layer slider, and pause highlighting completes the workflow. This tool significantly streamlines the prototyping workflow and empowers users to leverage the full capabilities of their manufacturing hardware, with future work directed at richer conditional anchoring, modal coordinate handling, and an expanded machine-command set.
